Analyzing Toronto Traffic Speeds: Understanding Urban Mobility, Safety, and Speed Dynamics

### 2024-12-03 paper/paper.qmd
finalize paper

---
title: "Analyzing Toronto Traffic Speeds: Insights into Urban Mobility, Safety, and Speed Dynamics"
subtitle: "Exploring Key Traffic Patterns and Relationships Between Speed Percentiles and Volume"
author: 
  - Andy Jiang
thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-vehicles-speed-analysis."
date: today
date-format: long
abstract: "This paper investigates traffic speed data from Toronto collected between 2017 and 2024, focusing on trends in speed percentiles and their relationship to traffic volume. Using a Bayesian linear regression model, we predict mean vehicle speeds based on key predictors such as 5th, 50th, and 95th percentile speeds and total traffic volume. Results highlight significant changes in traffic patterns post-2020, with increased extreme speeds linked to higher traffic volumes. This study provides insights into urban mobility dynamics, offering a basis for traffic management strategies and policy-making."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

Traffic speed analysis is a critical tool for understanding urban mobility, road safety, and congestion. In metropolitan areas like Toronto, shifting traffic patterns reflect changes in urban infrastructure, population growth, and behavioral adaptations to external events, such as the COVID-19 pandemic. The advent of technologies like radar-equipped speed signs has enabled the collection of detailed traffic data, providing new opportunities to analyze and predict traffic behavior at a granular level.

This paper focuses on traffic speed data collected in Toronto's designated Safety Zones between 2017 and 2024. The dataset comprises monthly summaries of speed percentiles, vehicle counts across specific speed ranges, and total traffic volumes. Notably, the study explores trends in extreme speeds (vehicles exceeding 100 km/h) and evaluates how they have evolved before and after 2020, a period marked by rapid population growth and post-pandemic recovery. These analyses aim to address a key gap: understanding how percentile-based speed measures and traffic volume influence overall mean speeds.

@@

The remainder of this paper is structured as follows. @sec-data discusses the raw data, cleaning process, and key variables of interest, offering visual and tabular representations of the dataset. @sec-model introduces and justifies the Bayesian linear regression model used to analyze the relationship between traffic volume, speed percentiles, and mean vehicle speed. @sec-results presents the findings of the model, highlighting significant trends and correlations within the data. Finally, @sec-discussion explores the broader implications of these results, offering insights into traffic behavior and potential avenues for future research. Additional information regarding data observation methodology is provided in [Appendix -@sec-dataobservation], and further model diagnostics are detailed in [Appendix -@sec-model-details].

@@

The dataset aggregates metrics monthly, offering insights into how traffic behavior has changed over time. This analysis is particularly focused on extreme speeds (vehicles exceeding 100 km/h) and their relationship with total traffic volume, as well as trends in key speed percentiles.

@@

However, the dataset has certain limitations. The number of vehicles recorded by the speed signs is not equivalent to true traffic volume, as it may exclude vehicles that pass too quickly or fall outside the radar’s range. Additionally, the presence of the signs might influence driver behavior, leading to changes in speed as drivers approach them. Furthermore, the data is limited to Safety Zones where the speed signs have been installed, restricting its spatial coverage. Despite these limitations, the dataset is well-documented, complete, and up-to-date, providing valuable insights into traffic speed patterns across Toronto.

@@

The traffic speed dataset contains information collected from multiple locations and months, detailing percentile speeds, total traffic volume, and counts of vehicles traveling within specific speed ranges. The dataset aggregates these metrics monthly, offering insights into traffic behavior and variations over time.

@@

- Percentile speeds (5th, 10th, ... 95th) representing speed thresholds exceeded by corresponding proportions of vehicles.
- Traffic volume as the total number of vehicles observed monthly.
- Speed bins (e.g., 0-4 km/h, 5-9 km/h) for detailed breakdowns of vehicle counts across speed ranges.
- @tbl-speed-preview showcases a simplified preview of the dataset with selected columns to highlight key features.

```{r}
#| label: tbl-speed-preview
#| tbl-cap: Preview of analysis traffic speed dataset.
#| echo: false
#| tbl-pos: H

@@

@tbl-random-sample provides a random sample of 10 observations, offering an unbiased snapshot of key metrics such as speed percentiles, total traffic volume, and counts of vehicles exceeding 100 km/h. This selection illustrates the dataset's diversity without emphasizing specific patterns.

```{r}
#| label: tbl-random-sample
#| tbl-cap: Random sample of traffic speed dataset.
#| echo: false
#| tbl-pos: H

@@

Vehicles traveling at extreme speeds (over 100 km/h) represent a critical factor in understanding traffic safety and violations. @tbl-extreme-speeds highlights the top 5 months with the highest counts of vehicles exceeding 100 km/h. This provides insights into when extreme speeds are most prevalent.

@@

Another key insight from this dataset is the proportion of vehicles traveling within a moderate speed range (50–70 km/h). This measure helps understand how typical driving speeds align with safe and efficient traffic flow. @tbl-speed-proportion highlights monthly proportions of vehicles in this speed range.

@@

```{r}
#| label: fig-key-percentiles
#| fig-cap: Trends in key speed percentiles (5th, 50th, and 95th) and the spread between the 95th and 5th percentiles over time.
#| echo: false
#| warning: false
#| message: false
# Summarize Key Percentiles and Calculate Speed Spread
key_percentiles <- data %>%
  group_by(month) %>%
  summarise(
    pct_05 = mean(pct_05, na.rm = TRUE),
    pct_50 = mean(pct_50, na.rm = TRUE),
    pct_95 = mean(pct_95, na.rm = TRUE)
  ) %>%
  mutate(speed_spread = pct_95 - pct_05)  # Calculate spread

@@

The above model was chosen to investigate the relationship between the average speed (`spd_mean`) of vehicles and key predictors, including total traffic volume (`volume`) and speed percentiles (`pct_05`, `pct_50`, and `pct_95`). These variables were selected because they provide a comprehensive view of traffic patterns and vehicle behavior across different speed thresholds.

The percentiles represent critical thresholds of speed distribution, with the 5th percentile (`pct_05`) capturing the slowest vehicles, the 50th percentile (`pct_50`) representing the median speed, and the 95th percentile (`pct_95`) reflecting the fastest vehicles. These thresholds are expected to influence the average speed significantly. Similarly, total traffic volume (`volume`) accounts for overall road congestion and is anticipated to impact the mean speed inversely, as higher volumes may reduce average speeds due to congestion.

@@

The Bayesian linear regression model results, summarized in @tbl-modelsummary, provide insight into the relationship between mean traffic speed (`spd_mean`) and key predictors such as speed percentiles and traffic volume.

```{r}
#| label: tbl-modelsummary
#| tbl-cap: Model summary of the predicted impact of traffic volume and speed percentiles on mean speed.
#| echo: false
#| warning: false
#| message: false
#| tbl-pos: H

modelsummary(mean_speed_model)
```

@@

These results highlight the dominant role of speed percentiles in determining mean speed while raising questions about the negligible impact of traffic volume. Further exploration into potential interactions or omitted variables could clarify this finding.

@@

While the Bayesian linear regression model employed in this paper provides meaningful insights into the relationship between key predictors and mean vehicle speed, several limitations warrant discussion. As highlighted in @sec-model, the model does not account for potential interactions among predictors, such as the interplay between traffic volume and percentile speeds. For instance, higher traffic volumes may disproportionately affect the 5th percentile speed compared to the 95th percentile, but this relationship remains unexplored due to the model’s linear structure.

@@

Future modeling efforts could incorporate interaction terms and hierarchical structures to better capture the complexity of traffic dynamics. Similarly, integrating data from diverse monitoring systems (e.g., highway cameras or GPS datasets) could provide a more comprehensive view of Toronto’s traffic behavior across different contexts.

@@

These delays underscore broader challenges in traffic enforcement, including bureaucratic hurdles, resource constraints, and reliance on traditional enforcement methods. Moreover, the cultural perspective within traffic policing, which often deprioritizes ticketing for non-criminal violations, exacerbates these issues. However, these challenges also present opportunities for leveraging technology to complement traditional enforcement and create a more comprehensive traffic management framework.

@@

The findings in this study offer a foundation for improving traffic safety and efficiency in Toronto. A central takeaway is the value of automated enforcement in addressing critical challenges, including extreme speeds, variability in traffic patterns, and enforcement gaps. Automated speed enforcement (ASE) cameras have already demonstrated significant success in Toronto, reducing both the prevalence of speeding and overall vehicle speeds in monitored areas. Expanding the use of these cameras and integrating new functionalities, such as detecting transit lane violations or intersection blockages, can further enhance their impact. Learning from other cities like Washington, D.C., and New York, Toronto can adopt similar innovations to target a broader range of traffic offenses [@Elliott_2024].

Urban road design remains another critical area for intervention. Consistent median speeds and limited speed variability are essential for maintaining safe and efficient traffic flow. To achieve this, urban planning efforts should focus on designing roads that naturally discourage dangerous speeds, while avoiding features that create bottlenecks or increase risks for vulnerable road users. For instance, implementing physical design elements such as raised crosswalks or lane-narrowing strategies can complement automated enforcement measures.

@@

The observational data analyzed in this study stems from the City of Toronto’s Safety Zone Watch Your Speed Program (WYSP), where radar-equipped speed signs monitor and aggregate vehicle speeds. While this data provides valuable insights, it also introduces limitations such as restricted spatial and temporal coverage, potential behavioral biases due to the visibility of speed signs, and an inability to capture all traffic volume accurately. This section proposes an idealized methodology to address these limitations and enhance the robustness of traffic speed data collection.

@@

- Temporal Sampling
  - Data collection should account for temporal variability by monitoring during different times of day (e.g., peak vs. off-peak hours), days of the week (e.g., weekdays vs. weekends), and seasons. This approach ensures a comprehensive understanding of traffic patterns and their fluctuations.

- Technological Enhancements
  - Advanced technologies such as LiDAR and AI-based sensors can supplement traditional radar devices to increase measurement accuracy. These technologies can detect multiple vehicles simultaneously, reducing blind spots and capturing speeds across a wider field of view.

@@

- Cost Implications: Deploying advanced devices and rotational monitoring systems requires significant investment in equipment and personnel.
- Coordination with Stakeholders: Accessing certain road segments for device installation may require coordination with local authorities, businesses, and residents.
- Data Management: Large-scale data collection introduces challenges related to storage, integration, and processing. Ensuring data integrity and consistency is critical for effective analysis.

Despite these challenges, the proposed framework offers a robust and realistic solution for improving traffic data collection. By addressing the limitations of the existing WYSP dataset, this methodology provides a foundation for more accurate and actionable insights into urban mobility.

@@

An evaluation of three years of data from Toronto's Automated Speed Enforcement (ASE) cameras provides valuable insights into the potential of automated systems to influence driver behavior and enhance traffic safety. Conducted by researchers from SickKids Hospital and Toronto Metropolitan University, this study highlights the effectiveness of ASE in reducing speeding across multiple speed zones [@Rodrigues_2023].

@@

- Rotational Deployment: Cameras were rotated across locations to broaden coverage and minimize behavioral biases from static monitoring.
- Comprehensive Data Collection: Despite some limitations imposed by the COVID-19 pandemic, the study captured substantial data during periods of normal activity, enhancing its reliability.
- Focused Interventions: Placement of ASE cameras in high-risk areas, such as school zones, demonstrated a targeted approach to improving road safety.

@@

The WYSP dataset offers valuable observational insights into Toronto’s traffic patterns, but it is not without its limitations.

@@

- Ease of Collection: Radar speed signs provide a non-intrusive method for collecting large amounts of data over extended periods.
- Focus on Safety Zones: Data collection in designated Safety Zones highlights areas with specific traffic and safety concerns.
- Detailed Metrics: The dataset includes percentile-based speed metrics (e.g., 5th, 50th, and 95th percentiles), which are critical for analyzing traffic behavior.

>>>>

---
title: "Analyzing Toronto Traffic Speeds: Understanding Urban Mobility, Safety, and Speed Dynamics"
subtitle: "Exploring Traffic Patterns and Relationships Between Speed Percentiles and Volume"
author: 
  - Andy Jiang
thanks: "Code and data are available at: https://github.com/AndyYanxunJiang/toronto-vehicles-speed-analysis."
date: today
date-format: long
abstract: "This paper investigates traffic speed data from Toronto collected between 2017 and 2024, focusing on trends in speed percentiles and their relationship to traffic volume. Using a Bayesian linear regression model, we predict mean vehicle speeds based on key predictors such as 5th, 50th, and 95th percentile speeds and total traffic volume. Results highlight significant changes in traffic patterns post-2020, with increased extreme speeds linked to higher traffic volumes. This study provides a deeper understanding of urban mobility dynamics, supporting traffic management strategies and policy-making."
format: pdf
number-sections: true
bibliography: references.bib
toc: true
---

@@

Traffic speed analysis is an essential method for understanding urban mobility, road safety, and congestion. In metropolitan areas like Toronto, shifting traffic patterns reflect changes in urban infrastructure, population growth, and behavioral adaptations to external events, such as the COVID-19 pandemic. The advent of technologies like radar-equipped speed signs has enabled the collection of detailed traffic data, providing new opportunities to analyze and predict traffic behavior at a granular level.

This paper focuses on traffic speed data collected in Toronto's designated Safety Zones between 2017 and 2024. The dataset comprises monthly summaries of speed percentiles, vehicle counts across specific speed ranges, and total traffic volumes. Notably, the study examines trends in extreme speeds (vehicles exceeding 100 km/h) and evaluates how they have changed before and after 2020, a period marked by rapid population growth and post-pandemic recovery. These analyses aim to address a key gap: understanding how percentile-based speed measures and traffic volume influence overall mean speeds.

@@

The remainder of this paper is structured as follows. @sec-data discusses the raw data, cleaning process, and key variables of interest, offering visual and tabular representations of the dataset. @sec-model introduces and justifies the Bayesian linear regression model used to analyze the relationship between traffic volume, speed percentiles, and mean vehicle speed. @sec-results presents the findings of the model, highlighting significant trends and correlations within the data. Finally, @sec-discussion explores the broader implications of these results, offering understandings of traffic behavior and potential avenues for future research. Additional information regarding data observation methodology is provided in [Appendix -@sec-dataobservation], and further model diagnostics are detailed in [Appendix -@sec-model-details].

@@

The dataset aggregates metrics monthly, offering understandings of how traffic behavior has changed over time. This analysis is particularly focused on extreme speeds (vehicles exceeding 100 km/h) and their relationship with total traffic volume, as well as trends in key speed percentiles.

@@

However, the dataset has certain limitations. The number of vehicles recorded by the speed signs is not equivalent to true traffic volume, as it may exclude vehicles that pass too quickly or fall outside the radar’s range. Additionally, the presence of the signs might influence driver behavior, leading to changes in speed as drivers approach them. Furthermore, the data is limited to Safety Zones where the speed signs have been installed, restricting its spatial coverage. Despite these limitations, the dataset is well-documented, complete, and up-to-date, providing useful understandings of traffic speed patterns across Toronto.

@@

The traffic speed dataset contains information collected from multiple locations and months, detailing percentile speeds, total traffic volume, and counts of vehicles traveling within specific speed ranges. The dataset aggregates these metrics monthly, offering understandings of traffic behavior and variations over time.

@@

- Percentile speeds (5th, 10th, ... 95th) representing speed thresholds exceeded by corresponding proportions of vehicles.
- Traffic volume as the total number of vehicles observed monthly.
- Speed bins (e.g., 0-4 km/h, 5-9 km/h) for detailed breakdowns of vehicle counts across speed ranges.

@tbl-speed-preview showcases a simplified preview of the dataset with selected columns to highlight key features.

```{r}
#| label: tbl-speed-preview
#| tbl-cap: Simplified preview of analysis traffic speed dataset.
#| echo: false
#| tbl-pos: H

@@

@tbl-random-sample provides a simplified random sample of 10 observations, offering an unbiased snapshot of key metrics such as speed percentiles, total traffic volume, and counts of vehicles exceeding 100 km/h. This selection illustrates the dataset's diversity without emphasizing specific patterns.

```{r}
#| label: tbl-random-sample
#| tbl-cap: Simplified random sample of analysis traffic speed dataset.
#| echo: false
#| tbl-pos: H

@@

Vehicles traveling at extreme speeds (over 100 km/h) represent a significant factor in understanding traffic safety and violations. @tbl-extreme-speeds highlights the top 5 months with the highest counts of vehicles exceeding 100 km/h. This provides understandings of when extreme speeds are most prevalent.

@@

Another perspective from this dataset is the proportion of vehicles traveling within a moderate speed range (50–70 km/h). This measure helps understand how typical driving speeds align with safe and efficient traffic flow. @tbl-speed-proportion highlights monthly proportions of vehicles in this speed range.

@@

```{r}
#| label: fig-key-percentiles
#| fig-cap: Trends in key speed percentiles (5th, 50th, and 95th) and the spread between the 95th and 5th percentiles over time.
#| echo: false
#| warning: false
#| message: false

# Summarize Key Percentiles and Calculate Speed Spread
key_percentiles <- data %>%
  group_by(month) %>%
  summarise(
    pct_05 = mean(pct_05, na.rm = TRUE),
    pct_50 = mean(pct_50, na.rm = TRUE),
    pct_95 = mean(pct_95, na.rm = TRUE)
  ) %>%
  mutate(speed_spread = pct_95 - pct_05)  # Calculate spread

@@

The above model was chosen to investigate the relationship between the average speed (`spd_mean`) of vehicles and key predictors, including total traffic volume (`volume`) and speed percentiles (`pct_05`, `pct_50`, and `pct_95`). These variables were selected because they provide a complete view of traffic patterns and vehicle behavior across different speed thresholds.

The percentiles represent significant thresholds of speed distribution, with the 5th percentile (`pct_05`) capturing the slowest vehicles, the 50th percentile (`pct_50`) representing the median speed, and the 95th percentile (`pct_95`) reflecting the fastest vehicles. These thresholds are expected to influence the average speed significantly. Similarly, total traffic volume (`volume`) accounts for overall road congestion and is anticipated to impact the mean speed inversely, as higher volumes may reduce average speeds due to congestion.

@@

The Bayesian linear regression model results, summarized in @tbl-modelsummary, provide understandings of the relationship between mean traffic speed (`spd_mean`) and key predictors such as speed percentiles and traffic volume.

```{r}
#| label: tbl-modelsummary
#| tbl-cap: Model summary of the predicted impact of traffic volume and speed percentiles on mean speed.
#| echo: false
#| warning: false
#| message: false
#| fig-pos: H

modelsummary(list("Mean Speed Prediction" = mean_speed_model))
```

@@

These results highlight the dominant role of speed percentiles in determining mean speed while raising questions about the negligible impact of traffic volume. Further examination into potential interactions or omitted variables could clarify this finding.

@@

While the Bayesian linear regression model employed in this paper provides meaningful interpretations of the relationship between key predictors and mean vehicle speed, several limitations warrant discussion. As highlighted in @sec-model, the model does not account for potential interactions among predictors, such as the interplay between traffic volume and percentile speeds. For instance, higher traffic volumes may disproportionately affect the 5th percentile speed compared to the 95th percentile, but this relationship remains unexplored due to the model’s linear structure.

@@

Future modeling efforts could incorporate interaction terms and hierarchical structures to better capture the complexity of traffic dynamics. Similarly, integrating data from diverse monitoring systems (e.g., highway cameras or GPS datasets) could provide a more complete view of Toronto’s traffic behavior across different contexts.

@@

These delays underscore broader challenges in traffic enforcement, including bureaucratic hurdles, resource constraints, and reliance on traditional enforcement methods. Moreover, the cultural perspective within traffic policing, which often deprioritizes ticketing for non-criminal violations, exacerbates these issues. However, these challenges also present opportunities for leveraging technology to complement traditional enforcement and create a more complete traffic management framework.

@@

The findings in this study offer a foundation for improving traffic safety and efficiency in Toronto. A central takeaway is the value of automated enforcement in addressing significant challenges, including extreme speeds, variability in traffic patterns, and enforcement gaps. Automated speed enforcement (ASE) cameras have already demonstrated significant success in Toronto, reducing both the prevalence of speeding and overall vehicle speeds in monitored areas. Expanding the use of these cameras and integrating new functionalities, such as detecting transit lane violations or intersection blockages, can further enhance their impact. Learning from other cities like Washington, D.C., and New York, Toronto can adopt similar innovations to target a broader range of traffic offenses [@Elliott_2024].

Urban road design remains another significant area for intervention. Consistent median speeds and limited speed variability are essential for maintaining safe and efficient traffic flow. To achieve this, urban planning efforts should focus on designing roads that naturally discourage dangerous speeds, while avoiding features that create bottlenecks or increase risks for vulnerable road users. For instance, implementing physical design elements such as raised crosswalks or lane-narrowing strategies can complement automated enforcement measures.

@@

The observational data analyzed in this study stems from the City of Toronto’s Safety Zone Watch Your Speed Program (WYSP), where radar-equipped speed signs monitor and aggregate vehicle speeds. While this data provides useful understandings, it also introduces limitations such as restricted spatial and temporal coverage, potential behavioral biases due to the visibility of speed signs, and an inability to capture all traffic volume accurately. This section proposes an idealized methodology to address these limitations and enhance the robustness of traffic speed data collection.

@@

- Temporal Sampling
  - Data collection should account for temporal variability by monitoring during different times of day (e.g., peak vs. off-peak hours), days of the week (e.g., weekdays vs. weekends), and seasons. This approach ensures a complete understanding of traffic patterns and their fluctuations.

- Technological Enhancements
  - Technologies such as LiDAR and AI-based sensors can supplement traditional radar devices to increase measurement accuracy. These technologies can detect multiple vehicles simultaneously, reducing blind spots and capturing speeds across a wider field of view.

@@

- Cost Implications: Deploying devices and rotational monitoring systems requires significant investment in equipment and personnel.
- Coordination with Stakeholders: Accessing certain road segments for device installation may require coordination with local authorities, businesses, and residents.
- Data Management: Large-scale data collection introduces challenges related to storage, integration, and processing. Ensuring data integrity and consistency is significant for effective analysis.

Despite these challenges, the proposed framework offers a robust and realistic solution for improving traffic data collection. By addressing the limitations of the existing WYSP dataset, this methodology provides a foundation for more accurate and actionable understandings of urban mobility.

@@

An evaluation of three years of data from Toronto's Automated Speed Enforcement (ASE) cameras provides useful understandings of the potential of automated systems to influence driver behavior and enhance traffic safety. Conducted by researchers from SickKids Hospital and Toronto Metropolitan University, this study highlights the effectiveness of ASE in reducing speeding across multiple speed zones [@Rodrigues_2023].

@@

- Rotational Deployment: Cameras were rotated across locations to broaden coverage and minimize behavioral biases from static monitoring.
- Extensive Data Collection: Despite some limitations imposed by the COVID-19 pandemic, the study captured substantial data during periods of normal activity, enhancing its reliability.
- Focused Interventions: Placement of ASE cameras in high-risk areas, such as school zones, demonstrated a targeted approach to improving road safety.

@@

The WYSP dataset offers useful observational understandings into Toronto’s traffic patterns, but it is not without its limitations.

@@

- Ease of Collection: Radar speed signs provide a non-intrusive method for collecting large amounts of data over extended periods.
- Focus on Safety Zones: Data collection in designated Safety Zones highlights areas with specific traffic and safety concerns.
- Detailed Metrics: The dataset includes percentile-based speed metrics (e.g., 5th, 50th, and 95th percentiles), which are significant for analyzing traffic behavior.
